\ttitle

### 2013-01-31 Chapters/Chapter2.tex
More on pca

\subsection{Principal Component Analysis}
When working with simple 2- or 3-dimensional data it is often, for humans, 
easy to discover patterns in the set. The data can be plotted and the lines or 
planes among which the data points lie gives an indication of the pattern. 
With Principal Component Analysis (PCA) this process is also easy for 
automated systems. With this method the overall form of a set of data points 
can be represented, the most discriminative dimensions can be found, the 
dimensionality of a set can be reduced (which yield in data compression) or 
the result can be used to characterize and differentiate sets of data points.

The workings of the PCA \cite{smith2002tutorial} relies on the concepts of 
standard deviation, 
covariance, eigenvectors and eigenvalues of matrices. These concepts will be 
discussed very briefly. The standard deviation of a set is an measure for the 
spread around the mean for a single dimension, or feature. The covariance 
between two features indicates how they are related; when positive the two 
features will increase together; when negative one will decrease when the 
other increases and when zero they are unrelated.
*** [eigenvectors]
*** [eigenvalues]

>>>>

\subsection{Principal Component Analysis}
When working with simple 2- or 3-dimensional data it is often, for humans, 
easy to discover patterns in the set. The data can be plotted and the lines or 
planes among which the data points lie gives an indication of the pattern. 
With Principal Component Analysis (PCA) this process is also possible for 
automated systems. With this method the overall form of a set of data points 
can be represented, the most discriminative dimensions can be found, the 
dimensionality of a set can be reduced (which yield in data compression) or 
the result can be used to characterize and differentiate sets of data points.

The workings of the PCA \cite{smith2002tutorial} relies on the concepts of 
standard deviation, 
covariance, eigenvectors and eigenvalues of matrices. These concepts will be 
discussed very briefly \footnote{For a more in-depth discussion we would like 
to refer the reader to \cite{jolliffe2005principal}}. The standard deviation 
of a set is an measure for the 
spread around the mean for a single dimension, or feature. The covariance 
between two features (dimensions of the data points) indicates how they are 
related; when positive the two 
features will increase together; when negative one will decrease when the 
other increases and when zero they are unrelated. The covariance matrix gives 
all the covariances for all pairs of features. This matrix is symmetrical 
about the diagonal and the values on the diagonal are the variances for each 
feature. For this matrix the eigenvectors can be computed. Eigenvectors have 
the characteristic that when they multiply a matrix, the resulting vector is a 
multiple of the eigenvector. The amount by which it is the multiple is the 
Eigenvalue. This is illustrated in formulae \ref{eq:no-eigenvector} and 
\ref{eq:eigenvector}. The second formula shows an Eigenvector and its 
associated Eigenvalue, 4. The vector in the first formula is not an 
Eigenvector.

\begin{equation} \label{eq:no-eigenvector}
	\begin{pmatrix} 2 & 3 \\ 2 & 1 \end{pmatrix}
\times
	\begin{pmatrix} 1 \\ 3 \end{pmatrix}
=
	\begin{pmatrix} 2 \cdot 1 + 3 \cdot 3 \\ 2 \cdot 1 + 1 \cdot 3 
	\end{pmatrix}
=
	\begin{pmatrix} 11 \\ 5 \end{pmatrix}
\end{equation}

\begin{equation} \label{eq:eigenvector}
	\begin{pmatrix} 2 & 3 \\ 2 & 1 \end{pmatrix}
\times
	\begin{pmatrix} 3 \\ 2 \end{pmatrix}
=
	\begin{pmatrix} 2 \cdot 3 + 3 \cdot 2 \\ 2 \cdot 3 + 1 \cdot 2  
	\end{pmatrix}
=
	\begin{pmatrix} 12 \\ 8 \end{pmatrix}
=
	4 \times \begin{pmatrix} 3 \\ 2 \end{pmatrix}
\end{equation}

It are these Eigenvectors and Eigenvalues that makes PCA possible. Given a 
matrix of $n \times n$ (only square matrices have Eigenvectors), $n$ 
Eigenvectors can be found. Each Eigenvector is perpendicular, or orthogonal, 
to the others. The Eigenvectors combined describes the lines over which the 
data is plotted. The Eigenvector with the largest Eigenvalue is the 
\emph{principal} component of the data set. It is said that it is the most 
significant dimension. Thus the Eigenvectors, and thereby the features or 
dimensions of the data set, can be sorted on significance by the Eigenvalues.

With this ordering there are a few applications possible. The first is to just 
make a (comprehensible) representation of the data. The Eigenvectors describe 
the cloud of data points and thus a

*** [graph, a bit of formulae with sets and dimensions]
